Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-05-03 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Edits to WGBS bioinformatics analysis

\begin{chapabstract}
I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used \emph{methylC-seq} protocol. All data used in my thesis were generated using this protocol.

@@

The chapter concludes by introducing the `methtuple` software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci. \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
\end{chapabstract}

@@

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLL}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabulary}
\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
\label{tab:bs-seq_mappers}
\end{table}

@@

There is generally no way to tell if a read is truly a PCR duplicated based on the sequencing data alone.

There is generally no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify suspected PCR duplicates (which are almost always inaccurately referred to as 'PCR duplicates'[^pcr_dup]). Software to identify PCR duplicates includes the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}. These software all use mapped reads from a `SAM` or `BAM` file as input.

@@

Roughly speaking, these tools will flag reads with identical start and end co-ordinates as being suspected PCR duplicates. This will inevitably lead to some false positives because reads may have identical co-ordinates yet not be PCR duplicates. The false positive rate is a particular problem when a subset of the genome is sequenced at high coverage, such as in RRBS. This can be thought of as an example of the pigeonhole principle, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).

@@

In practice, this means that for whole-genome sequencing data we can be fairly confident that suspected PCR duplicates are _true_ PCR duplicates. However, for targetted sequencing, such as RRBS or amplicon sequencing, we are much less confident and may remove some of our signal if we remove possible PCR duplicates. Instead, for RRBS we might exclude regions with an "abnormally" high sequencing coverage \citep{Krueger:2012ks}. For amplicon sequencing we often can't afford to exclude possible PCR duplicates if, for example, the aim is to identify rare epialleles by very deep sequencing of a small region.

@@

1. Compute the M-bias from the aligned reads, then call methylation events. The methylation calling should include filters to remove the detected M-bias (along with any other additional filters). This strategy requires two passes over the `SAM/BAM` file - one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}.
2. Call methylation events but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the file containing the methylation calls. This is the approach taken by `BSmooth` \citep{Hansen:2012gr}.

I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple`, described in Section \ref{sec:methtuple}).

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption[\emph{E18VA} M-bias]{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately. For CpGs in \emph{read-1}, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in \emph{read-1}. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\label{fig:E18VA_M-bias}
\end{figure}

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e. read-positions whose methylation levels are 'too far away' from the vast majority of read-positions. It is rather a subjective decision to make. Two problems that I found when making these decisions were:

@@

This strategy is generally sufficient because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads. However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101 in 150 bp reads. Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants to exclude[^methtuple_mbias].

In the datasets I have analysed I've had to ignore up to 30 read-positions per read due to M-bias. Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely. One idea is to inversely weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

Prior to methylation calling, each read should be filtered to remove low-quality reads and low quality bases. When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters]. Strictly speaking each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads. In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.

@@

The gold-standard is to perform whole-genome DNA sequencing of each sample. This data is then used to form a set of sample-specific methylation loci. However, this approach is also very expensive due to the extra sequencing requirements. A cheaper alternative is to genotype the sample on a genome-wide SNP microarray. This will give very accurate, very cheap genotypes at a large number of loci ($500,000$ to $5,000,000$). However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.

@@

The third approach, and arguably the bare minimum, is to call methylation loci using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population. For example, we might exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.

@@

This approach can obviously only exclude sites of known variation from consideration, and cannot add sample-specific methylation loci. To remove those reference-specific methylation loci that are __not__ found in databases of known genetic variation, we might identify loci in the sample that display a large number of non-C/T bases (resp. non-G/A bases) at a C (resp. G) on the forward (resp. reverse) strand of the reference genome.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci, that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns.}
\label{fig:incomplete_picture_1-tuples}
\end{figure}

@@

Generally, when referring to m-tuples I implicitly mean those with $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple. The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1} $--$ \texttt{5}). The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}

@@

With current sequencing technology we are limited to extracting m-tuples that span no more than $200-250$ bp. This obviously affects the size of m-tuples that we can study. Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:datasets} for a description of the Lister dataset). Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples. This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples. Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a 'spike' can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for over 50 whole-genome bisulfite-sequencing samples. Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).

@@

The maximum memory usage is not strictly proportional to chromosome length. It is instead driven by the number and density of CpGs on the chromosome.  For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Table \ref{tab:cpg_density_hg19}). The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain. Therefore, memory usage is relatively constant across values of `-m` for a given chromosome. The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`. This means that all 2-tuples with $NIL \geq 0$ were computed and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_memory}
\end{figure}

@@

__TODO: necessary?__

>>>>

\begin{chapabstract}
I explain the bioinformatics analysis of whole-genome bisulfite-sequencing data, concentrating on the most widely used methylC-seq protocol. All data used in my thesis were generated using this protocol.

@@

The chapter concludes by introducing the \texttt{methtuple} software, a unique methylation caller for extracting methylation patterns at tuples of methylation loci. \texttt{methtuple} is critical for work in later chapters on co-methylation (Chapter \ref{chap:co-methylation}) but has wider application in facilitating downstream analyses of bisulfite-sequencing data.
\end{chapabstract}

@@

\begin{table}[h]
\centering
\caption[Popular bisulfite-sequencing read mappers and their underlying mapping software]{Four popular bisulfite-sequencing read mappers, selected to highlight the variety of underlying mapping software used by these tools.}
\label{tab:bs-seq_mappers}
\begin{tabulary}{\textwidth}{LLL}
\toprule
Name      & Reference                                            & Underlying mapping software \\ \midrule
Bismark   & \citet{Krueger:2011eb}                              & Bowtie1 or Bowtie2          \\
bwa-meth  & \citet{Pedersen:2014wt}                             & bwa-mem                     \\
BSMAP     & \citet{Xi:2009bd}                                   & SOAP                        \\
Novoalign & \url{http://www.novocraft.com/products/novoalign/} & Novoalign                   \\ \bottomrule
\end{tabulary}
\end{table}

@@

There is generally no way to tell based on sequencing data if a read is truly a PCR duplicate. However, it is relatively easy to identify suspected PCR duplicates (which are almost always referred to as 'PCR duplicates'[^pcr_dup]). Software to identify PCR duplicates includes the `MarkDuplicates` function that is a part of the `Picard` software ([http://broadinstitute.github.io/picard/](http://broadinstitute.github.io/picard/)), the `rmdup` function that is a part of the `SAMtools` software \citep{Li:2009ka} and `SAMBLASTER` \citep{Faust:2014hf}. These software all use mapped reads from a `SAM` or `BAM` file as input.

@@

Roughly speaking, these tools will flag reads with identical start and end co-ordinates as being suspected PCR duplicates. This will inevitably lead to some false positives because reads may have identical co-ordinates yet not be true PCR duplicates. The false positive rate is a particular problem when a subset of the genome is sequenced at high coverage, such as in RRBS. This can be thought of as an example of the pigeonhole principle, which states that if we have $m$ containers (positions where a read can start) and $n > m$ items (reads), then at least one container must contain more than one item (at least one position must have more than one read starting there).

@@

In practice, for whole-genome sequencing data we can be fairly confident that suspected PCR duplicates are _true_ PCR duplicates. However, for targetted sequencing, such as RRBS or amplicon sequencing, we are much less confident and may remove some of our signal if we remove possible PCR duplicates. Instead, for RRBS we might exclude regions with an "abnormally" high sequencing coverage \citep{Krueger:2012ks}. For amplicon sequencing we often can't afford to exclude possible PCR duplicates if, for example, the aim is to identify rare epialleles by very deep sequencing of a small region.

@@

1. Compute the M-bias from the aligned reads, then call methylation events. The methylation calling should include filters to remove the detected M-bias (along with any other additional filters). This strategy requires two passes over the `SAM/BAM` file, one to compute the M-bias and one to do the methylation calling. This is the approach used by `bismark_methylation_extractor` \citep{Krueger:2011eb} and `Bis-SNP` \citep{Liu:2012ge}.
2. Call methylation events but retain the read-position of each methylation event. Compute the M-bias from this first file and then filter out methylation events that suffer from M-bias. This strategy requires only a single pass over the `SAM/BAM` file but requires additional information to be stored alongside the methylation calls which is then followed by a pass over the file containing the methylation calls. This is the approach taken by `BSmooth` \citep{Hansen:2012gr}.

I find the first strategy conceptually simpler, and easier to program, so use it in my methylation calling software, `methtuple` (described in Section \ref{sec:methtuple}).

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption[\emph{E18VA} M-bias]{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of \emph{read-1} (R1) and \emph{read-2} (R2) are plotted separately. For CpGs in \emph{read-1}, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in \emph{read-2}, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read-positions for all methylation types. The spike at read-position 101 is also evident, albeit to a lesser extent, in \emph{read-1}. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\label{fig:E18VA_M-bias}
\end{figure}

@@

In practice, the M-bias curves for each sample are visually inspected to look for evidence of M-bias, i.e. read-positions whose methylation levels are 'too far away' from the majority of read-positions. It is rather a subjective decision to make. Two problems that I found when making these decisions were:

@@

This strategy is generally sufficient because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads. However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101 in 150 bp reads. Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. The `methtuple` software avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants to exclude (discussed in Section \ref{sec:methtuple}).

In the datasets I have analysed, I have had to ignore up to 30 read-positions per read due to M-bias. Ideally, we would be able to remove the effects of M-bias by accounting for it in methylation calling rather than by simply excluding those read-positions entirely. One idea is to inversely weight methylation calls by the level of M-bias. A downside to this approach is that this would turn an otherwise binary methylation call into a continuous value between $0$ and $1$, with an attendant loss in interpretability. However, we could still compute the ubiquitous $\beta$-values, traditionally defined as the proportion of reads that are methylated at a locus, and use these in downstream inferences, without much loss in interpretation. The bigger problem with this approach is the increased computational complexity and cost, and this is why I have not further pursued this idea.

@@

Prior to methylation calling, each read should be filtered to remove low-quality reads and low quality bases. When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters]. Strictly speaking, each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads. In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.

@@

The gold-standard is to perform whole-genome DNA sequencing of each sample. This data is then used to form a set of sample-specific methylation loci. This approach, however, is also very expensive due to the extra sequencing requirements. A cheaper alternative is to genotype the sample on a genome-wide SNP microarray. This will give very accurate, very cheap genotypes at a large number of loci ($500,000$ to $5,000,000$). However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.

@@

The third approach, and arguably the bare minimum, is to call methylation events using the reference genome and to _post-hoc_ exclude any loci that overlap sites of known genetic variation in the population. For example, we might exclude all cytosines in the reference genome that are also SNPs in dbSNP \citep{Sherry:2001gh}.

@@

This approach can obviously only exclude sites of known variation from consideration, and cannot add sample-specific methylation loci. To remove those reference-specific methylation loci that are __not__ found in databases of known genetic variation, we might identify loci in the sample that display a large number of non-C/T bases (resp. non-G/A bases) at a C (resp. G) on the forward (resp. reverse) strand of the reference genome[^ad_hoc_bissnp]

[^ad_hoc_bissnp]: This is like an _ad hoc_ and limited version of `Bis-SNP`.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/incomplete_picture_1-tuples.pdf}
\caption[Different methylation patterns can lead to identical $\beta$-values]{Two regions, each with four methylation loci that have identical $\beta$-values ($\beta = \frac{M}{M + U}$) at 1-tuples yet have very different overall methylation patterns. Each line is a read, a white circle is an unmethylated CpG and a black circle is a methylated CpG.}
\label{fig:incomplete_picture_1-tuples}
\end{figure}

@@

When referring to m-tuples I implicitly mean those with $NIL = 0$; I will explicitly use the notation $NIL \geq 0$ when I wish to make clear that there may be intervening methylation loci in the m-tuple. The default option of `methtuple` is to produce m-tuples with $NIL = 0$ unless the `--all-combinations` flag is set[^nil0].

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption[\texttt{methtuple} example region]{Diagram of three-paired end reads (\texttt{A}, \texttt{B} and \texttt{C}) mapping to a region containing five methylation loci (\texttt{1}, \texttt{2}, \texttt{3}, \texttt{4} and  \texttt{5}). The suffix \texttt{\_1} or \texttt{\_2} indicates whether it is \emph{read-1} or \emph{read-2}, respectively.}
\label{fig:methtuple_example1}
\end{figure}

@@

With current sequencing technology we are limited to extracting m-tuples that span no more than $200$ to $250$ bp. This obviously affects the size of m-tuples that we can study. Figure \ref{fig:Lister_CpGs_per_read} shows the number of CpGs per read for the Lister dataset (see Chapter \ref{chap:datasets} for a description of the Lister dataset). Longer reads, and paired-end reads, contain more methylation loci and so are more informative for analyses using m-tuples. This can be seen by comparing, for example, the _ADS_ and _HSF1_ samples. Samples sequenced more deeply will have more reads per m-tuple, although this can't be seen in these plots since they are normalised by sequencing depth.

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting these overlapping bases. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for $40$ whole-genome bisulfite-sequencing samples. Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).

@@

The maximum memory usage is not strictly proportional to chromosome length. It is instead driven by the number and density of CpGs on the chromosome.  For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Table \ref{tab:cpg_density_hg19}). The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain. Memory usage is therefore relatively constant across values of `-m` for a given chromosome. The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`. This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set.}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set.}
\label{fig:methtuple_memory}
\end{figure}

@@

This chapter has detailed the first steps in a bioinformatics analysis of whole-genome bisulfite-sequencing data. There are many decisions to be made at each step and these will ultimately affect the quality of the data available for downstream analysis.

This chapter also introduced `methtuple`, the first of several pieces of software that were developed as part of my thesis. `methtuple` is software for calling methylation patterns at m-tuples from whole-genome bisulfite-sequencing data. It will be essential for our later analysis of within-fragment co-methylation (Chapter \ref{chap:co-methylation}) and has wider utility in facilitating other downstream analyses based on methylation patterns at m-tuples (Chapter \ref{chap:wgbs_downstream_analyses}).
